Towards topological type theory for decrypting transfinite methods in classical mathematics

### 2025-08-25 slides-ai2025.tex
Beginning of AI slides

% compile using: agda --latex; result will then be in the latex/ subfolder
\documentclass[11pt,utf8,notheorems,compress,t,aspectratio=169]{beamer}
\usepackage{etex}

@@

\usepackage{mathtools}
\usepackage{xspace}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{manfnt}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{proof}
\usepackage{agda}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing,positioning,svg.path}
\hypersetup{colorlinks=true}
\usepackage[dvipsnames]{xcolor}
\usepackage{fontawesome5}

@@

\setbeameroption{show notes on second screen=bottom}
\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\footnotesize\justifying#2\par}}}
\makeatletter
\setbeamertemplate{note page}
{%
  {%
    \scriptsize
    \insertvrule{.25\paperheight}{white!90!black}
    \vskip-.25\paperheight
    \nointerlineskip
    \vbox{
      %\hfill\insertslideintonotes{0.25}\hskip-\Gm@rmargin\hskip0pt%
      %\vskip-0.25\paperheight%
      %\nointerlineskip
      \begin{pgfpicture}{0cm}{0cm}{0cm}{0cm}
        \begin{pgflowlevelscope}{\pgftransformrotate{90}}
          {\pgftransformshift{\pgfpoint{-2cm}{0.2cm}}%
          \pgftext[base,left]{\footnotesize\the\year-\ifnum\month<10\relax0\fi\the\month-\ifnum\day<10\relax0\fi\the\day}}
        \end{pgflowlevelscope}
      \end{pgfpicture}}
    \nointerlineskip
    \vbox to .25\paperheight{\vskip0.5em
      \hbox{\insertshorttitle[width=8cm]}%
      \setbox\beamer@tempbox=\hbox{\insertsection}%
      \hbox{\ifdim\wd\beamer@tempbox>1pt{\hskip4pt\raise3pt\hbox{\vrule
            width0.4pt height7pt\vrule width 9pt
            height0.4pt}}\hskip1pt\hbox{\begin{minipage}[t]{7.5cm}\def\breakhere{}\insertsection\end{minipage}}\fi%
      }%
      \setbox\beamer@tempbox=\hbox{\insertsubsection}%
      \hbox{\ifdim\wd\beamer@tempbox>1pt{\hskip17.4pt\raise3pt\hbox{\vrule
            width0.4pt height7pt\vrule width 9pt
            height0.4pt}}\hskip1pt\hbox{\begin{minipage}[t]{7.5cm}\def\breakhere{}\insertsubsection\end{minipage}}\fi%
      }%
      \setbox\beamer@tempbox=\hbox{\insertshortframetitle}%
      \hbox{\ifdim\wd\beamer@tempbox>1pt{\hskip30.8pt\raise3pt\hbox{\vrule
            width0.4pt height7pt\vrule width 9pt
            height0.4pt}}\hskip1pt\hbox{\insertshortframetitle[width=7cm]}\fi%
      }%
      \vfil}%
  }%
  \vskip.25em
  \nointerlineskip
  \insertnote
}
\makeatother

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\vspace*{-7em}\includegraphics[width=\paperwidth]{forest2}\end{minipage}}
\begin{frame}[c]
  \centering
  \color{white}

  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip

@@

  \setbeamercolor{block body}{bg=black!100}
  \begin{minipage}{0.52\textwidth}
    \begin{block}{}
      \centering\normalsize\color{white}
      \hil{\color{white}Towards topological type theory for decrypting
      transfinite methods in classical~mathematics} \\[-0.4em]
      \
    \end{block}
  \end{minipage}

  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bf
  \colorbox{black}{\begin{minipage}{0.2\textwidth}
    \centering
    TYPES 2025 \\
    June 9th, 2025
  \end{minipage}}
  \bigskip

  \colorbox{black}{\begin{minipage}{0.2\textwidth}
    \centering
    Ingo Blechschmidt \\
    University of Antwerp
  \end{minipage}}
\end{frame}}

@@

\addtocounter{framenumber}{-1}
\begin{frame}\frametitle{A classical logic fairy tale}
  \centering\bigskip\bigskip\bigskip
  \includegraphics[height=9.5em]{philosophers-stone}

  \jnote{1}{
    \vspace*{-1em}
    \begin{changemargin}{-1.5em}{-1.5em}
      \setlength{\columnsep}{15pt}
      \begin{multicols}{2}
        \justifying
        \scriptsize

        \textbf{Narrator.} Once upon a time, in a kingdom far, far away, the
        queen of the country and of all Möbius strips called for her royal
        philosopher.

        \textbf{Queen.} Philosopher! I ask you to carry out the following order.
        Get me the Philosopher's Stone, or alternatively find out how one could
        produce arbitrary amounts of gold with it!

        \textbf{Philosopher.} But my queen! I haven't studied anything useful!
        How could I fulfill this order?

        \textbf{Queen.} That is not my concern. I'll see you again tomorrow.
        Should you not accomplish the task, I will take your head off.

        \textbf{Narrator.} After a long and wakeful night the philosopher was
        called to the queen again.

        \textbf{Queen.} Tell me! What do you have to report?

        \textbf{Philosopher.} It was not easy and I needed to follow lots of
        obscure references, but finally I actually found out how to use the
        Philosopher's Stone to produce arbitrary amounts of gold. But only I can
        conduct this procedure, your royal highness.

        \textbf{Queen.}
        Alright. So be it.

        \textbf{Narrator.} And so years passed by, during which the philosopher
        imagined herself to be safe. The queen searched for the stone on her own,
        but as long as she hadn't found it, the philosopher didn't need to worry.
        Yet one day the impossible happened: The queen has found the stone! And
        promptly called for her philosopher.

        \textbf{Queen.} Philosopher, look! I have found the Philosopher's Stone!
        Now live up to your promise! \emph{[She hands over the stone.]}

        \textbf{Philosopher.} Thank you. \emph{[She inspects the stone.]} This is
        indeed the Philosopher's Stone. Many years ago you asked me to either
        acquire the Philosopher's Stone or find out how to produce arbitrary
        amounts of gold using it. Now it's my pleasure to present to you the
        Philosopher's Stone. \emph{[She returns the stone.]}
      \end{multicols}
    \end{changemargin}
  }

  \jnote{2}{
    Adapted from
    \href{http://blog.ezyang.com/2013/04/a-classical-logic-fairy-tale/}{Edward
    Yang's blog}.

    The idea conveyed in the fairy tale is at the core of the
    \emph{constructive} proof of $\neg\neg(A \vee \neg A)$. While the law of
    excluded middle (\textsc{lem}) is not available in constructive
    mathematics, up to a double negation every instance of it is.

    This observation gives rise to the double negation translation, a
    transformation of (statements and) proofs in classical logic to proofs in
    intuitionistic logic, at the only expense of introducing double negations;
    by following this up with Friedman's trick with the nontrivial exit
    continuation (``Baby Barr Theorem''), we can in many cases even get rid of
    these double negations in a second step.

    Getting rid of applications of the law of excluded middle is in many
    contexts (an amazing mind-boggingly wonder at first, but then) a routine
    matter---see \href{https://rt.quasicoherent.io/}{here} for more details and
    pointers to the literature.
  }
\end{frame}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{1.7em}\includegraphics[height=\paperheight]{sea-of-clouds-1}\end{minipage}}
\begin{frame}{A case study in Hilbert's program}
  \jnote{1}{
    Well quasi-orders are an important notion in proof theory and termination
    analysis.
  }

  \jnote{2-3}{
    \vspace*{-0.5em}
    The presented proof rests on the law of excluded middle and hence cannot
    immediately be interpreted as a program for finding suitable indices~$i <
    j$. However, constructive proofs are also possible (for instance by
    induction on the value of a given term of the sequence, see
    \href{https://www.math.lmu.de/~petrakis/Dickson.pdf}{Constructive
    combinatorics of Dickson's Lemma} by Iosif Petrakis for several fine
    quantitative results). And even more:
    There is a procedure for regarding this proof---and many others in the
    theory of well quasi-orders---as \emph{blueprints} for more informative
    constructive proofs. This shall be our motto for today:

    \emph{Do not take classical proofs literally, instead ask which
    constructive proofs they are blueprints for.}
  }

  \jnote{3}{
    The displayed stability results, along with several others, provide a
    flexible toolbox for constructing new well quasi-orders from given ones.
    However, with the classical formulation of \emph{well},
    shortly to be renamed~``well\sinf'', these results are
    \href{https://lets-play-agda.quasicoherent.io/Padova2025.Explorations.Forcing.Orthogonal.html}{inherently
    classical} (even the weaker claim ``the product of two streamless sets is
    streamless'' admits countermodels).

    (In Higman's lemma, the set~$X^*$ of finite lists of elements of~$X$ is
    equipped with the following ordering: We have~$(x_0 \dblcolon \ldots
    \dblcolon x_{n-1} \dblcolon [\,]) \leq (y_0 \dblcolon \ldots \dblcolon
    y_{m-1} \dblcolon [\,])$ iff there is an increasing injection~$f : \{
    0,\ldots,n-1 \} \to \{ 0,\ldots,m-1 \}$ such that~$x_i \leq y_{f(i)}$ for all~$i < n$.
    Kruskal's theorem is the result behind the celebrated enormous number
    $\mathsf{TREE}(3)$.)
  }

  \jnote{7}{
    Thanks to work by Thierry Coquand, Daniel Fridlender and Monika
    Seisenberger, a constructive substitute is available, the notion well\ind.
    In classical mathematics (where bar induction is available), this notion is
    equivalent to well\sinf.

    The predicate~$\mathsf{Good}$ of finite lists~$\sigma = (x_0 \dblcolon
    \ldots \dblcolon x_{n-1} \dblcolon [\,])$ appearing in the definition
    of~$G$ expresses that some earlier term is at most some later term,
    i.\,e.\@ that there merely exist~$i < j$ such that~$x_i \leq x_j$.

    The operator $\dblcolon^r$ is adding an element at the end.
  }

  \jnote{8}{
    \begin{columns}[t]
      \begin{column}{0.65\textwidth}
        The original notion well\sinf:
        \begin{itemize}
          \justifying
          \item[\cmark] short and simple
          \\[-1.3em]
          \item[\cmark] constructively satisfied for the main examples (but
          only because of the theory around well\ind)
          \\[-1.3em]
          \item[\cmark] concise abstract proofs (albeit employing transfinite methods)
          \\[-1.3em]
          \item[\xmark] main results not constructively attainable
          \\[-1.3em]
          \item[\xmark] philosophically strenuous by the quantification over all sequences
          \\[-1.3em]
          \item[\xmark] not stable under ``change of base''---a forcing extension
          of the universe may well contain more sequences than the base universe
          \\[-1.3em]
          \item[\xmark] negative (universal) condition
        \end{itemize}
      \end{column}

      \begin{column}{0.33\textwidth}
        The constructive substitute well\ind:
        \begin{itemize}
          \justifying
          \item[\cmark] main results constructive
          \\[-1.3em]
          \item[\cmark] stable under change of base
          \\[-1.3em]
          \item[\cmark] positive (existential) condition
          \\[-1.3em]
          \item[\xmark] proofs intriguing, but also somewhat alien, not just some
          trivial reshuffling of the classical arguments; classical sequence
          language cannot be used
        \end{itemize}
      \end{column}
    \end{columns}
  }

  \textbf{Def.} Let~$(X,{\leq})$ be a quasi-order.
  \vspace*{-0.3em}
  \begin{itemize}
    \item A sequence~$\alpha : \NN \to X$ is \hil{good} iff
    there merely exist~$i < j$ with~$\alpha\,i \leq \alpha\,j$.
    \item The quasi-order~$X$ is \hil{well\only<4->{\sinf}} iff every sequence~$\NN \to X$ is good.
  \end{itemize}
  \pause
  \vspace*{-1.3em}

  \begin{columns}[t]
    \begin{column}{0.46\textwidth}
      \begin{block}{Natural numbers\phantom{y}}
        \textbf{Prop.} $(\NN, {\leq})$ is well\only<4->{\sinf}. \formalized
        \smallskip

        \emph{Proof.} Let~$\alpha : \NN \to \NN$. By~\badbox{\textsc{lem}}, there is a
        \bad{minimum}~$\alpha\,i$.
        Set~$j \defeq i + 1$. \qed

        \hfill\color{gray}offensive?
      \end{block}
      \only<2-6>{
        \vspace*{-2em}
        \[ \astikznodetransparentlycircled{xm}{7}\!,
        \quad \astikznodetransparentlycircled{x0}{4}\!,
        \quad \astikznodecircled{t1}{mypurple}{3}\!,
        \quad \astikznodetransparentlycircled{x1}{1}\!,
        \quad \astikznodecircled{t2}{mypurple}{8}\!,
        \quad \astikznodetransparentlycircled{x2}{2}\!,
        \quad \ldots \]
      {\centering\begin{tikzpicture}[remember picture,overlay]
        \node[draw=mypurple, circle, thick, inner sep=0.1em] (t3) {\scriptsize$\leq$};
        \path[draw=mypurple,thick]
          (t1)
          to [out=-90, in=180] (t3)
          to [out=0, in=-90] (t2);
      \end{tikzpicture}\par}}
    \end{column}

    \pause

    \begin{column}{0.50\textwidth}
      \begin{block}{Key stability results}
        \justifying
        \only<1-5>{Assuming~\badbox{\textsc{lem}}
        and~\badbox{\textsc{dc}}}\only<6->{\good{Constructively}}, \ldots

        \hil{Dickson:} \tabto{1.7cm} If~$X$ and~$Y$ are well\only<4-5>{\sinf}\only<6->{\ind}, so is~$X \times Y$. \\
        \hil{Higman:}  \tabto{1.7cm} If~$X$ is well\only<4-5>{\sinf}\only<6->{\ind}, so is~$\mathrm{List}\,X$. \\
        \hil{Kruskal:} \tabto{1.7cm} If~$X$ is well\only<4-5>{\sinf}\only<6->{\ind}, so is~$\mathrm{Tree}\,X$.
      \end{block}
    \end{column}
  \end{columns}
  \pause
  \pause
  \pause
  \pause

  \textbf{Def.} A quasi-order~$X$ is \hil{well\ind} iff~$G\,[\,]$, where $G$ is
  the following inductively defined predicate on \hil{finite lists}.~\formalized{}
  (In presence of \bad{bar induction}, $\text{well}\ind \Leftrightarrow
  \text{well}\sinf$.)
  \[
    \small
    \infer{\mathsf{now}\,p : G\,\sigma}{p : \mathsf{Good}\,\sigma}
    \qquad
    \infer{\mathsf{later}\,f : G\,\sigma}{f : (x : X) \to G\,(\sigma \dblcolon^r x)}
  \]
  \vspace*{-1.3em}

  \only<8>{
    \begin{columns}[c]
      \begin{column}{0.01\textwidth}
	\includegraphics[height=2.4em]{question-mark}
      \end{column}
      \begin{column}{0.9\textwidth}
      Is there a procedure for reinterpreting \hil{classical proofs}
      regarding well\sinf{} as \newline \hil{blueprints for constructive proofs} regarding well\ind?
      \end{column}
    \end{columns}
  }
\end{frame}}

\begin{frame}{Missing functions in the type of all functions?}
  \jnote{1}{
    The double negation translation technique fails to constructivize the two
    proofs presented on the slide.

    The reason is that, after applying the translation, the ``$I$-extracted
    subsequence'' appearing in the proof on the left will no longer actually be
    a function~$\NN \to X$. Instead, it will be a partially defined function
    with the property that each number~$n$ does \emph{not~not} belong to the
    domain. However, the assumption that~$X$ is well\sinf{} does not apply to
    these kinds of generalized sequences.

    In contrast, the notion well\ind{} is much stronger. If~$X$ is well\ind,
    then not only will every actual sequence~$\NN \to X$ be good, but (in a
    suitable sense) so will every up-to-$\neg\neg$-partially-defined sequence,
    every multivalued sequence and many more kinds of generalized sequences.
  }

  \begin{columns}[t]
    \begin{column}{0.57\textwidth}
      \begin{block}{Behold: A transfinite tool \ldots\phantom{g}}
        \justifying
        \textbf{Lemma.} \badbox{\textsc{lem}} Let~$X$ be well\sinf.
        Let~$\alpha : \NN \to X$. Then there is an increasing
        subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.
      \end{block}

      \small
      \emph{Proof.}
      \begin{changemargin}{-0.8em}{0em}\begin{enumerate}\justifying
      \item The type
      $I \defeq \sum_{i \? \NN} \neg\sum_{j \? \NN} i < j \times \alpha\,i \leq
      \alpha\,j$
      is not in bijection with~$\NN$, as else the~$I$-extracted subsequence
      of~$\alpha$ would not be good.\smallskip
      \item By~\badbox{\textsc{lem}}, the type~$I$ is finite.\smallskip
      \item Any index~$i_0$ larger than all the indices in~$I$ is a suitable
      starting point for an increasing subsequence. \qed
      %  \item Let $\mathsf{Sad}\,i \defeq
      %  \neg \mathsf{Happy}\,i$ be the proposition
      %  expressing that~$i$ does not appear as the first component
      %  of a good pair, where
      %  $\mathsf{Happy}\,i \defeq \sum_{j \? \NN} i < j \times \alpha\,i \leq
      %  \alpha\,j$.
      %  \item If~$\mathsf{Sad}$ is anonymously dense in~$\NN$, i.e. $\forall a
      %  \? \NN\_ \neg\neg\exists b\?\NN\_ b \geq a \times \mathsf{Sad}\,b$, then
      %  \badbox{\textsc{lem}} and \badbox{\textsc{dc}} provide us with a
      %  subsequence~$\alpha\,k_0 \leq \alpha\,k_1 \leq \ldots$
      %  with~$\mathsf{Sad}\,k_0, \mathsf{Sad}\,k_1, \ldots$. This is a contradiction to wellness.
      %  \item
      %  Hence there is~\bad{not~not} an index~$a \? \NN$ such that~$\mathsf{Sad}$ is
      %  strictly bounded by~$a$; hence every index~$b \geq a$ does~\bad{not not} have
      %  a partner: $\forall b\geq a\_ \neg\neg\exists c>b\_
      %  \alpha\,b \leq \alpha\,c$. Thus~\bad{\textsc{lem}} and \bad{\textsc{dc}}
      %  provide us with the desired subsequence. \qed
      \end{enumerate}\end{changemargin}
    \end{column}

    \begin{column}{0.43\textwidth}
      \begin{block}{\ldots{} implying a concrete consequence}
        \textbf{Cor.} \badbox{\textsc{lem}}
        Let~$X$ and~$Y$ be well\sinf. Then~$X \times Y$ is
        well\sinf.\phantom{g}
      \end{block}

      \small
      \emph{Proof.}
      \begin{changemargin}{-0.8em}{0em}\begin{enumerate}\justifying
      \item Let a sequence~$\gamma : \NN \to X \times Y$ be
      given. Write~$\gamma\,k = (\alpha\,k,\beta\,k)$.\smallskip
      \item By the lemma, there is an
      increasing subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.\smallskip
      \item Because~$Y$
      is well, there are indices~$n < m$ such that~$\beta\,i_n \leq
      \beta\,i_m$.\smallskip
      \item As
      also~$\alpha\,i_n \leq \alpha\,i_m$, the sequence~$\gamma$ is good. \qed
      \end{enumerate}\end{changemargin}
    \end{column}
  \end{columns}

  \dnote{\emph{We cannot trust~\bad{\textsc{lem}}-provided sequences to be
  available in the type~$\NN \to X$. Similarly with~\bad{\textsc{dc}}.}}
\end{frame}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{0.29cm}\hspace*{-1cm}\includegraphics[height=\paperheight]{tower-cropped}\end{minipage}}
\begin{frame}{Where do many cherished inductive definitions come from?}
  \begin{changemargin}{8.0em}{-0.5em}
    In mathematics, we routinely enlarge structures:
    \begin{itemize}
      \small
      \item Pass from~$\mathbb{Q}$ to~$\mathbb{R}$, to embrace irrationals.
      \item Pass from~$\mathbb{R}$ to~$\mathbb{C}$, to obtain $\sqrt{-1}$.
      \item Pass from~$\mathbb{C}$ to~$\mathbb{C}[X]$, to obtain a ``generic number''.
    \end{itemize}

    In set and type theory, we can also enlarge \hil{the mathematical universe}:
    \begin{itemize}
      \small
      \item Force a \hil{generic sequence} $\NN \to X$.
            \fixedhref{https://lets-play-agda.quasicoherent.io/Padova2025.Explorations.Forcing.Cohen.html}{\formalized}
      \item Force a \hil{generic enumeration} $\NN \twoheadrightarrow X$
            (even if $X$ is uncountable).
            \fixedhref{https://lets-play-agda.quasicoherent.io/Padova2025.Explorations.Forcing.Levy.html}{\formalized}
      \item Force a \hil{generic prime ideal} of a given ring.
            \fixedhref{https://iblech.github.io/constructive-maximal-ideals/Forcing.Prime.html}{\formalized}
    \end{itemize}
    \pause

    Central observations of the multiversal yoga:
    \begin{itemize}
      \small
      \item A quasi-order~$X$ is well\ind iff the generic sequence~$\NN \to X$
      is good.
      \item A set is Noetherian iff the generic sequence~$\NN \to X$
      has two repeated terms.
      \item A relation is well-founded iff for the generic descending chain,
      $\bot$.
      \item A ring element is nilpotent iff it is contained in the generic
      prime ideal.
      \fixedhref{https://iblech.github.io/constructive-maximal-ideals/Forcing.Prime.html\#corollary}{\formalized}
    \end{itemize}
    \pause

    The mystery of nongeometric sequents:
    \begin{itemize}
    \small
      \item The generic ring is a field.
      \item \ \\[-1.2em]\mbox{For the generic surjection $f : \NN \twoheadrightarrow \RR$,
      $\neg\neg \exists n \? \NN\_ f(n) = \pi \wedge f(n+1) = e$.
      \fixedhref{https://lets-play-agda.quasicoherent.io/Padova2025.Explorations.Forcing.Levy.html}{\formalized}}
    \end{itemize}
  \end{changemargin}

  \jnote{1}{
    The technique for enlarging the mathematical universe is called
    \emph{forcing} and was originally pioneered by Paul Cohen. Set theorists
    use forcing to construct new models of set theory from given ones, in order
    to explore the range of set-theoretic possibility. For instance, by forcing
    we can construct models of~\textsc{zfc} validating the continuum hypothesis
    and also models which falsify it.

    We here refer to a simplification of original forcing which is useful in a
    constructive metatheory. At its core, every forcing extension is just a
    formula and proof translation of a certain form. For instance, there is a
    forcing extension validating~\bad{\textsc{lem}} even if the base universe
    does not; this forcing extension is not a deep mystery, for a statement holds
    in that forcing extension iff its double negation translation holds in the
    base universe and it is well-known that the double negation translation
    of every instance of~\bad{\textsc{lem}} is an intuitionistic tautology.

    \href{https://www.speicherleck.de/iblech/stuff/slides-herrsching2023.pdf}{Here
    is a set of slides on constructive forcing}, and Section~4 of
    \href{https://raw.githubusercontent.com/iblech/constructive-maximal-ideals/master/tex/extended.pdf}{this
    joint paper with Peter Schuster} contains a written summary of constructive
    forcing.
  }

  \jnote{2}{
    Forcing the existence of a new function~$\NN \to X$ is also called
    \emph{Cohen forcing}. Forcing the existence of a new surjection is called
    \emph{Lévy collapse}. Forcing the existence of a prime ideal does not have
    an established name, perhaps it should be called \emph{Zariski forcing}; it
    has been explored by Thierry Coquand and others of the constructive algebra
    crew.

    The multiversal yoga allows us to reason with the good notion well\ind{}
    using the simple language of well\sinf, and similarly with several other
    notions.
  }

  \jnote{3}{
    For more on the mystery of nongeometric sequents, see
    \href{https://www.speicherleck.de/iblech/stuff/slides-birmingham2019.pdf}{this
    set of slides}.
  }
\end{frame}}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
\begin{frame}{A modal language for harnessing the multiverse}
  \jnote{1}{
    By \emph{topos}, we mean \emph{Grothendieck topos}. In constructive
    forcing, a ``forcing extension of the base universe'' is exactly the same
    thing as a Grothendieck topos.

    A particular member of the rich and varied landscape of toposes is the
    \emph{trivial topos}, in which every statement whatsoever holds. By
    restricting to positive toposes, we exclude this special case.

>>>>

\documentclass[12pt,utf8,notheorems,compress,t,aspectratio=169]{beamer}
\usepackage{etex}

@@

\usepackage{mathtools}
\usepackage{xspace}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{manfnt}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{environ}
\usepackage{proof}
\usepackage{agda}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing,positioning,svg.path}
\hypersetup{colorlinks=true}
\usepackage[dvipsnames]{xcolor}
\usepackage{fontawesome5}

@@

%\setbeameroption{show notes on second screen=bottom}
\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\footnotesize\justifying#2\par}}}
\makeatletter
\setbeamertemplate{note page}
{%
  {%
    \scriptsize
    \insertvrule{.25\paperheight}{white!90!black}
    \vskip-.25\paperheight
    \nointerlineskip
    \vbox{
      %\hfill\insertslideintonotes{0.25}\hskip-\Gm@rmargin\hskip0pt%
      %\vskip-0.25\paperheight%
      %\nointerlineskip
      \begin{pgfpicture}{0cm}{0cm}{0cm}{0cm}
        \begin{pgflowlevelscope}{\pgftransformrotate{90}}
          {\pgftransformshift{\pgfpoint{-2cm}{0.2cm}}%
          \pgftext[base,left]{\footnotesize\the\year-\ifnum\month<10\relax0\fi\the\month-\ifnum\day<10\relax0\fi\the\day}}
        \end{pgflowlevelscope}
      \end{pgfpicture}}
    \nointerlineskip
    \vbox to .25\paperheight{\vskip0.5em
      \hbox{\insertshorttitle[width=8cm]}%
      \setbox\beamer@tempbox=\hbox{\insertsection}%
      \hbox{\ifdim\wd\beamer@tempbox>1pt{\hskip4pt\raise3pt\hbox{\vrule
            width0.4pt height7pt\vrule width 9pt
            height0.4pt}}\hskip1pt\hbox{\begin{minipage}[t]{7.5cm}\def\breakhere{}\insertsection\end{minipage}}\fi%
      }%
      \setbox\beamer@tempbox=\hbox{\insertsubsection}%
      \hbox{\ifdim\wd\beamer@tempbox>1pt{\hskip17.4pt\raise3pt\hbox{\vrule
            width0.4pt height7pt\vrule width 9pt
            height0.4pt}}\hskip1pt\hbox{\begin{minipage}[t]{7.5cm}\def\breakhere{}\insertsubsection\end{minipage}}\fi%
      }%
      \setbox\beamer@tempbox=\hbox{\insertshortframetitle}%
      \hbox{\ifdim\wd\beamer@tempbox>1pt{\hskip30.8pt\raise3pt\hbox{\vrule
            width0.4pt height7pt\vrule width 9pt
            height0.4pt}}\hskip1pt\hbox{\insertshortframetitle[width=7cm]}\fi%
      }%
      \vfil}%
  }%
  \vskip.25em
  \nointerlineskip
  \insertnote
}
\makeatother

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\vspace*{-0em}\includegraphics[width=\paperwidth]{lost-melody-2}\end{minipage}}
\begin{frame}[c]
  \centering
  \color{white}

  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip

@@

  \setbeamercolor{block body}{bg=black!100}
  \begin{minipage}{0.40\textwidth}
    \begin{block}{}
      \centering\normalsize\color{white}
      \hil{\color{white} Proofs as programs?} \\[-0.4em]
      \
    \end{block}
  \end{minipage}

  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bf
  \colorbox{black}{\begin{minipage}{0.3\textwidth}
    \centering
    AI Transforms Math Research \\
    University of Augsburg \\
    August 26th, 2025
  \end{minipage}}
  \bigskip

  \colorbox{black}{\begin{minipage}{0.3\textwidth}
    \centering
    Ingo Blechschmidt \\
    University of Antwerp
  \end{minipage}}
\end{frame}}

@@

\begin{frame}{In this talk}
  \bigskip
  \bigskip
  \begin{columns}
    \begin{column}{0.25\textwidth}
      \centering\footnotesize
      \includegraphics[height=2cm]{calabi-yau} \mbox{commutative algebra and}
      \\
      \mbox{number theory}
    \end{column}
    \begin{column}{0.2\textwidth}
      \centering\footnotesize
      \includegraphics[height=2cm]{turing-machine} \\ \mbox{numerical content}
    \end{column}
    \begin{column}{0.4\textheight}
      \centering\footnotesize
      \includegraphics[height=2cm]{3-adic-numbers} \\
      \mbox{a fractal} \mbox{without points}
    \end{column}
    \begin{column}{0.45\textheight}
      \centering\footnotesize
      \includegraphics[height=2cm]{cracks} \mbox{foundational crisis}
    \end{column}
  \end{columns}

  \medskip
  \medskip
  \medskip
  \medskip

  \begin{columns}
    \begin{column}{0.4\textheight}
      \centering\footnotesize
      \includegraphics[height=2cm]{multiverse} \mbox{traveling} \\
      the \mbox{multiverse}
    \end{column}
    \begin{column}{0.4\textheight}
      \centering\footnotesize
      \includegraphics[height=2cm]{monad} monadic side effects
    \end{column}
    \begin{column}{0.4\textheight}
      \centering\footnotesize
      \includegraphics[height=2cm]{hen-2} \\ proof assistants
    \end{column}
    \begin{column}{0.4\textheight}
      \centering\footnotesize
      \includegraphics[height=2cm]{alien} \\ alien algorithms
    \end{column}
  \end{columns}
\end{frame}

\NewEnviron{mining}[1][0.6\textwidth]{
  \includegraphics[height=2em,valign=c]{hammer-and-pick}\quad
  \colorbox{mypurple!10}{\begin{minipage}{#1}
    \BODY
  \end{minipage}}
}

\begin{frame}{A glimpse of proof mining}
  \textbf{Theorem.} For every natural number~$n$, there is a prime number
  larger than~$n$.~\fixedhref{https://unimath.github.io/agda-unimath/elementary-number-theory.infinitude-of-primes.html}{\formalized}

  \emph{Proof (Euclid).} Every prime factor of~$n! + 1$ will do. \qed
  \pause

  \begin{mining}
    Let~$p_0, p_1, p_2, \ldots$ be the sequence of prime numbers.

    \textbf{Scholium.} $p_{n+1} \leq p_n! + 1$.

    \textbf{Scholium.} $p_n \leq 2^{2^n}$.
  \end{mining}
  \pause
  \bigskip

  \emph{Proof (Euler).} If there were only finitely many primes,
  the identity
  \[ \prod_p \frac{1}{1-1/p} = \sum_{n \geq 1} \frac{1}{n} \]
  would imply that the harmonic series converges. \qed

  \begin{mining}
    \textbf{Scholium.} $p_n \leq \lceil e^{n+1-\gamma} \rceil$.
  \end{mining}
\end{frame}

% 1995: conjecture
% 2003: solution
% 2018: rate of convergence
\begin{frame}{Proof mining in convex analysis}
  \textbf{Theorem (zero displacement conjecture).}
  Let~$H$ be a Hilbert space. \newline
  Let~\mbox{$C_1, \ldots, C_N \subseteq H$} be nonempty
  closed convex subsets with orthogonal projections~$P_{C_i}$. Let~$T = P_{C_N}
  \circ \ldots \circ P_{C_1}$. Then for every~$x \in H$,
  \[ \| T^{n+1}x - T^nx \| \xrightarrow{n \to \infty} 0. \]

  \emph{Proof.} See
  \fixedhref{https://www.ams.org/journals/proc/2003-131-01/S0002-9939-02-06528-0/S0002-9939-02-06528-0.pdf}{[Bauschke
  2003]}, employing Minty's theorem, the Brézis--Haraux
  theorem, Rockafellar's maximal monotonicity and sum theorems, strongly
  nonexpansive mappings, conjugate functions, normal cones, \ldots \qed
  \pause

  \begin{mining}[0.85\textwidth]
    \textbf{Scholium
    \fixedhref{https://www2.mathematik.tu-darmstadt.de/~kohlenbach/inconsistentfeasibility.pdf}{[Kohlenbach
    2018]}.} In this situation, let~$b$ be an upper bound on the norm of~$x$ and let~$K$ be an
    upper bound for the norm of~$N$ arbitrary points $c_i \in C_i$. Then
    \[ \forall \varepsilon > 0\_
      \forall n \geq \phi(\varepsilon,N,b,K)\_
      \|T^{n+1}x - T^nx\| < \varepsilon, \]
    where~$\phi(\varepsilon,N,b,K)$ is given by a certain explicit formula.
  \end{mining}
\end{frame}

\begin{frame}{Proof mining in approximation theory}
  Let~$n \in \NN$. Let~$P_n$ be the space of polynomials of degree at most~$n$. \\
  Let~$f \in C[0,1]$ be a continuous function.

  \textbf{Theorem.} There is a \hil{unique} best~$L_1$-approximation of~$f$ in~$P_n$.
  \pause

  Let~$\omega$ be a \hil{modulus of uniform continuity} for~$f$,
  i.e. a function~$\mathbb{R}^+ \to \mathbb{R}^+$ such that
  \[ \forall x,\tilde x \in [0,1]\_ \forall \varepsilon > 0\_
    \bigl(|x-\tilde x| < \omega(\varepsilon) \Longrightarrow |f(x)-f(\tilde x)| <
    \varepsilon\bigr). \]

  \begin{mining}[0.85\textwidth]
    \textbf{Scholium
    \fixedhref{https://www2.mathematik.tu-darmstadt.de/~kohlenbach/Keele.pdf}{[Kohlenbach
    1990]}.} $\forall \varepsilon > 0\_ \forall p_1,p_2 \in P_n\_$
    \[
      \bigwedge_{i=1}^2 \bigl(\|f - p_i\|_1 - \operatorname{dist}_1(f, P_n) <
      \phi(\omega,n,\varepsilon)\bigr) \Longrightarrow \|p_1 - p_2\|_1 \leq
      \varepsilon, \]
    where~$\phi(\omega,n,\varepsilon)$ is given by a certain explicit formula.
  \end{mining}
\end{frame}
